Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-17 MS_thesis.tex
Change part I title to movivation and outline
Rephrase abstract

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis and encapsulated in fields such as population genetics, evolutionary game theory, and quantitative genetics. Historically, many of these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I use ideas from statistical physics to analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has therefore been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks. These SDEs also recover well-known results such as the replicator-mutator equation and Fisher’s fundamental theorem in the infinite population limit, and can help explain and organize several previous studies under a single theoretical framework. They also have implications for empirical studies and the formulation of evolutionary agent-based models. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations.
}

@@

%wrap the declaration of every new \part in the \clear and the group to ensure that the fresh page created by \part does not have a page number.
% This hack-ey workaround is from https://tex.stackexchange.com/a/191189 and is needed because \part defines \pagestyle WITHIN the definition itself and so you cannot use \thispagestyle to mess with it
\cleardoublepage
\begingroup
\makeatletter
\let\ps@plain\ps@empty
\part{Background}
\endgroup

@@

%You need the below commands to add a 'Part' level reference to the Appendices in the TOC
%from https://tex.stackexchange.com/a/286452
%The \phantomsection is needed for hyperref to jump to the correct location
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}

>>>>

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, many of these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit and can help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for evolutionary agent-based models and empirical studies.
}

@@

%wrap the declaration of every new \part in the \clear and the group to ensure that the fresh page created by \part does not have a page number.
% This hack-ey workaround is from https://tex.stackexchange.com/a/191189 and is needed because \part defines \pagestyle WITHIN the definition itself and so you cannot use \thispagestyle to mess with it
\cleardoublepage
\begingroup
\makeatletter
\let\ps@plain\ps@empty
\part{Motivation \& Outline}
\endgroup

@@

%The \phantomsection is needed for hyperref to jump to the correct location
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}
